Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-16 manuscript/manuscript.Rmd
updates ms with MAE statistics and more text on validation methods

Do we need detailed demographic data to forecast population responses to climate change? Probably.
========================================================================================

@@

We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models. We thinned these chains by 10 to reduce potential autocorrelation, leaving us with 300 samples for each parameter. At each time step, we randomly selected climate covariates from one of the 14 observed years. Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration. Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results. Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.

@@

### Model validation
To test each model's ability to forecast the population state we used leave-one-year-out cross validation. For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting. We repeated this procedure for all 13 observation years. This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level).

Results
-------
### Comparison of Forecast Models
We performed one-step-ahead forecasts to compare the two models (IPM vs. QBM). We did not use the random year effects for these forecasts. Thus, beyond the effect of genet size or quadrat cover, only the climate covariates could affect expect plant cover. The IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H. comata* and *P. smithii*; Table 1). In no case did the QBM significantly outperform the IPM (Table 1). As a metric of model uncertainty, we calculated the euclidean distance between the upper and lower 90% quantiles for each model's forecasts. The IPM always outperformed the QBM in this respect (Table 1), with consistently lower distances between the 90% quantiles. In general the IPM outperformed the QBM because it had (i) lower MAE for two species, (ii) statistically similar MAE for the other two species, and (iii) considerably more precise forecasts for all species.

### Forecasting Climate Change Impacts
Simulated equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Table 2). For the IPM, proportional changes range from about 1-25% for *B. gracilis*, 7-25% for *H. comata*, 8% for *P. smithii*, and 2-20% for *P. secunda*. For the QBM, proportional changes due to climate perturbation ranged from 20-100% for *B. gracilis*, 10-14% for *H. comata*, 0.25-8% for *P. smithii*, and 4-17% for *P. secunda*. However, the forecasts were extremely uncertain (Figure 1). The IPM produced more certain forecasts than the QBM, as expected based on the validation results (Table 1). But even for the IPM forecasts, predicted percent changes in median equilibrium cover ranged from -200% to 200% (Figure 1). Forecast uncertainty did not vary appreciably among species.

The population responses of species as forecast by the IPM is a function of the combined effects of individual vital rate sensitivities to climate and the contribution of each vital to population dynamics. For *B. gracilis*, all vital rates (survival, growth, and rectruitment) were positively affected by increases in precipitation and temperature (Figure 2). Intuitively, this leads to an overall increase in *B. gracilis* cover (Table 2).

Precipitation increase resulted in all vital rates contributing to an increase in simulated cover for *H. comata*. But there is a small negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK SIMULATIONS.

*P smithii* responded negatively to an increase in precipitation, temperature, or both regardless of the vital rate through which perturbed climate was acting. The effects were small, however, leading to relatively minor population consequences under climate change that did not vary much across scenarios (Table 2).

For *B. gracilis*, the IPM predicted a modest increase in cover with a 1% increase in the mean of precipitation or a 1% increase in the mean of temperature, the compounding effect of both being a 20% increase in cover. This reflects the relatively strong effects of precipitation and climate on *B. gracilis* genet growth and recruitment (Figure 1). The QBM also predicted increased *B. gracilis* cover with a precipitation increase, but increasing temperature decreased equilibrium cover (Figure 3).

The IPM and QBM produced consistent predictions for *H. comata* under increased precipitation and when both precipitation and temperature were increased (Figure 3). However, the IPM predicted more modest changes than the QBM, and for a temperature increase the two models differed: the IPM predicted an increase in cover while the QBM predicted the opposite.

@@

# mae_list <- list()
# mae_pvals <- list()
# species <- unique(qbmstats$species)
# for(spp in species){
#   tmp1 <- subset(tmpipm, species==spp)
#   tmp2 <- subset(tmpqbm, species==spp)
#   err1 <- as.data.frame(abs(tmp1$mean_obs - tmp1$mean_pred))
#   err2 <- as.data.frame(abs(tmp2$mean_obs - tmp2$mean_pred))
#   colnames(err1) <- "resid"
#   err1$model <- "ipm"
#   colnames(err2) <- "resid"
#   err2$model <- "qbm"
#   errd <-  rbind(err1, err2)
#   mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
#   mae_list[[spp]] <- mae_ttest
#   mae_pvals[[spp]] <- mae_ttest[["p.value"]]
# }
# 
# mae_ps <- melt(mae_pvals)

stats_table <- rbind(ipmstats, qbmstats)
stats_table <- stats_table[ order(stats_table[,"species"]), ]
stats_table <- stats_table[,c("species", "model", "mae",
                              "quant_dist","mean_cover")]
# comment          <- list()
# comment$pos      <- list()
# comment$pos[[1]] <- c(nrow(stats_table))
# comment$command  <- c(paste("\\hline \n",  # we`ll replace all default hlines with this and the ones below
#                             "The IPM MAE is significantly lower for HECO ($p = 3.3\times10^{-10}$) and POSE ($p = 0.00013$. MAEs are statisticially similar between models for BOGR and PASM.   \n",
#                             sep = ""))

colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
             include.rownames=FALSE, caption.placement="top")
```
The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$) and POSE ($p = 0.00013$). MAEs are statisticially similar between models for BOGR and PASM.

>>>>

Microscale models outperform mesoscale models when forecasting climate change impacts on plant populations
========================================================================================

@@

We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models. At each time step, we randomly selected climate covariates from one of the 14 observed years. Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration. Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results. Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.

@@

### Model validation
To test each model's ability to forecast the population state we used out of sample predictions using leave-one-year-out cross validation. For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting. Within each observation year, several quadrats are sampled. So we made predictions for each observed quadrat in the focal year initialized with cover the previous year. Since we were making quadrat specific predictions we incorporated the group effect on the intercept for both models. We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via randomdrawd from the MCMC chain as described above. Random year effects were set to zero since year effects cannot be assigned to unobserved years.

This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level). We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the difference between the observed cover for a given quadrat-year and the median prediction. To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors. We use MAE as our measure of accuracy. To measure precision calculated the distance between the upper and lower 90th quantiles of the 100 predictions as a measure of forecast precision and averaged this value over quadrat-years for each species.

### Testing sensitivity to climate covariates
Our main goal in this paper is to see if models based on aggregate level data are as sensitive to climate as models based on individual level data. So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%. We ran this simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in. Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain. We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model uncertainty on forecast precision.

As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM. Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, and recruitment vital rates individually and in pairwise combinations. This allows us to isolate the vital rate(s) most sensitive to climate. For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.

Results
-------
### Comparison of Forecast Models
The IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H. comata* and *P. smithii*; Table 1). In no case did the QBM significantly outperform the IPM (Table 1). The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1). In general the IPM outperformed the QBM because it had (i) lower MAE for two species, (ii) statistically similar MAE for the other two species, and (iii) considerably more precise forecasts for all species.

### Forecasting Climate Change Impacts

@@

mae_list <- list()
mae_pvals <- list()
species <- unique(qbmstats$species)
for(spp in species){
  tmp1 <- subset(avg_ipm, species==spp)
  tmp2 <- subset(avg_qbm, species==spp)
  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
  colnames(err1) <- "resid"
  err1$model <- "ipm"
  colnames(err2) <- "resid"
  err2$model <- "qbm"
  errd <-  rbind(err1, err2)
  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
  mae_list[[spp]] <- mae_ttest
  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
}

mae_ps <- melt(mae_pvals)

stats_table <- rbind(ipmstats, qbmstats)
stats_table <- stats_table[ order(stats_table[,"species"]), ]
stats_table <- stats_table[,c("species", "model", "mae",
                              "quant_dist","mean_cover")]

colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
print.xtable(xtable(stats_table, caption = "Accuracy (mean absoluet error, MAE) and precision (90\\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
             include.rownames=FALSE, caption.placement="top")
```
*NOTES*: The IPM MAE is significantly lower at $\alpha=0.05$ for *B. gracilis* (*P* = `r mae_ps[1,"value"]`), *H. comata* (*P* = `r mae_ps[2,"value"]`), and *P. smithii* (*P* = `r mae_ps[3,"value"]`). MAEs are statisticially similar between models for *P. secunda* (*P* = `r mae_ps[4,"value"]`).

### 2015-06-16 manuscript/manuscript.tex
updates ms with MAE statistics and more text on validation methods

\author{}
\date{}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing
\setlength\parindent{24pt}

@@

\section{Do we need detailed demographic data to forecast population
responses to climate change?
Probably.}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change-probably.}

@@

All R code and data necessary to replicate our analysis is archived on
GitHub as release 1.0
(\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
stable release will remain static as a record of this analysis, but
subsequent versions may appear if we update this work.

@@

To test each model's ability to forecast the population state we used
leave-one-year-out cross validation. For both levels of modeling, we fit
the vital rate models using observations from all years except one, and
then used those fitted parameters in the population models to perform a
one-step-ahead forecast for the year whose observations were withheld
from model fitting. We repeated this procedure for all 13 observation
years. This model validation allowed us to compare accuracy and
precision of the two modeling approaches (individual-level versus
population-level).

@@

We performed one-step-ahead forecasts to compare the two models (IPM
vs.~QBM). We did not use the random year effects for these forecasts.
Thus, beyond the effect of genet size or quadrat cover, only the climate
covariates could affect expect plant cover. The IPM had significantly
lower overall error (MAE, mean absolute error) for two species (\emph{H.
comata} and \emph{P. smithii}; Table 1). In no case did the QBM
significantly outperform the IPM (Table 1). As a metric of model
uncertainty, we calculated the euclidean distance between the upper and
lower 90\% quantiles for each model's forecasts. The IPM always
outperformed the QBM in this respect (Table 1), with consistently lower
distances between the 90\% quantiles. In general the IPM outperformed
the QBM because it had (i) lower MAE for two species, (ii) statistically
similar MAE for the other two species, and (iii) considerably more
precise forecasts for all species.

@@

Simulated equilibrium cover simulated from the models was sensitive to
climate perturbations, but the IPM and QBM produced inconsistent results
(Table 2). For the IPM, proportional changes range from about 1-25\% for
\emph{B. gracilis}, 7-25\% for \emph{H. comata}, 8\% for \emph{P.
smithii}, and 2-20\% for \emph{P. secunda}. For the QBM, proportional
changes due to climate perturbation ranged from 20-100\% for \emph{B.
gracilis}, 10-14\% for \emph{H. comata}, 0.25-8\% for \emph{P. smithii},
and 4-17\% for \emph{P. secunda}. However, the forecasts were extremely
uncertain (Figure 1). The IPM produced more certain forecasts than the
QBM, as expected based on the validation results (Table 1). But even for
the IPM forecasts, predicted percent changes in median equilibrium cover
ranged from -200\% to 200\% (Figure 1). Forecast uncertainty did not
vary appreciably among species.

The population responses of species as forecast by the IPM is a function
of the combined effects of individual vital rate sensitivities to
climate and the contribution of each vital to population dynamics. For
\emph{B. gracilis}, all vital rates (survival, growth, and rectruitment)
were positively affected by increases in precipitation and temperature
(Figure 2). Intuitively, this leads to an overall increase in \emph{B.
gracilis} cover (Table 2).

Precipitation increase resulted in all vital rates contributing to an
increase in simulated cover for \emph{H. comata}. But there is a small
negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK
SIMULATIONS.

\emph{P smithii} responded negatively to an increase in precipitation,
temperature, or both regardless of the vital rate through which
perturbed climate was acting. The effects were small, however, leading
to relatively minor population consequences under climate change that
did not vary much across scenarios (Table 2).

For \emph{B. gracilis}, the IPM predicted a modest increase in cover
with a 1\% increase in the mean of precipitation or a 1\% increase in
the mean of temperature, the compounding effect of both being a 20\%
increase in cover. This reflects the relatively strong effects of
precipitation and climate on \emph{B. gracilis} genet growth and
recruitment (Figure 1). The QBM also predicted increased \emph{B.
gracilis} cover with a precipitation increase, but increasing
temperature decreased equilibrium cover (Figure 3).

The IPM and QBM produced consistent predictions for \emph{H. comata}
under increased precipitation and when both precipitation and
temperature were increased (Figure 3). However, the IPM predicted more
modest changes than the QBM, and for a temperature increase the two
models differed: the IPM predicted an increase in cover while the QBM
predicted the opposite.

@@

\begin{table}[ht]
\centering
\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$)
and POSE ($p = 0.00013$). MAEs are statisticially similar between models
for BOGR and PASM.

@@

.

>>>>

\author{}
\date{}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\section{Microscale models outperform mesoscale models when forecasting
climate change impacts on plant
populations}\label{microscale-models-outperform-mesoscale-models-when-forecasting-climate-change-impacts-on-plant-populations}

@@

All R code and data necessary to reproduce our analysis is archived on
GitHub as release 1.0
(\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
stable release will remain static as a record of this analysis, but
subsequent versions may appear if we update this work.

@@

To test each model's ability to forecast the population state we used
out of sample predictions using leave-one-year-out cross validation. For
both levels of modeling, we fit the vital rate models using observations
from all years except one, and then used those fitted parameters in the
population models to perform a one-step-ahead forecast for the year
whose observations were withheld from model fitting. Within each
observation year, several quadrats are sampled. So we made predictions
for each observed quadrat in the focal year initialized with cover the
previous year. Since we were making quadrat specific predictions we
incorporated the group effect on the intercept for both models. We
repeated this procedure for all 13 observation years, making 100
one-step-ahead forecasts for each quadrat-year combination with
parameter uncertainty included via randomdrawd from the MCMC chain as
described above. Random year effects were set to zero since year effects
cannot be assigned to unobserved years.

This model validation allowed us to compare accuracy and precision of
the two modeling approaches (individual-level versus population-level).
We first calculated the median predicted cover across the 100
simulations for each quadrat-year and then calculated the absolute error
as the difference between the observed cover for a given quadrat-year
and the median prediction. To arrive at mean absolute error (MAE), we
then averaged the absolute error within each species across the
quadrat-year specific errors. We use MAE as our measure of accuracy. To
measure precision calculated the distance between the upper and lower
90th quantiles of the 100 predictions as a measure of forecast precision
and averaged this value over quadrat-years for each species.

\subsubsection{Testing sensitivity to climate
covariates}\label{testing-sensitivity-to-climate-covariates}

Our main goal in this paper is to see if models based on aggregate level
data are as sensitive to climate as models based on individual level
data. So, with our fitted and validated models in hand, we ran
simulations for each model type (IPM and QBM) under four climate
perturbation scenarios: (1) observed climate, (2) precipitation
increased by 1\%, (3) temperature increased by 1\%, and (4)
precipitation and temperature increased by 1\%. We ran this simulations
for 2,500 time steps, enough to estimate equilibrium cover after
discarding an initial 500 time steps as burn-in. Each simulation was run
under two parameter scenarios: (1) using mean parameter estimates and
(2) using randomly drawn parameters from the MCMC chain. We use (1) to
detect the overall sensitivity of equilibrium cover to climate, and we
use (2) to show the impact of model uncertainty on forecast precision.

As an effort to identify potential discrepencies between IPM and QBM
forecasts, we also ran simulations designed to quantify the
sensitivities of individual and combined vital rates to climate for the
IPM. Specifically, we ran simulations for the above climate scenarios,
but applied the perturbed climate covariates to survival, growth, and
recruitment vital rates individually and in pairwise combinations. This
allows us to isolate the vital rate(s) most sensitive to climate. For
this analysis, we used mean parameter estimates to reduce the sources of
uncertainty in the sensitivity estimates.

@@

The IPM had significantly lower overall error (MAE, mean absolute error)
for two species (\emph{H. comata} and \emph{P. smithii}; Table 1). In no
case did the QBM significantly outperform the IPM (Table 1). The IPM was
consistently more precise than the QBM, with lower distances between the
90\% quantiles across all species (Table 1). In general the IPM
outperformed the QBM because it had (i) lower MAE for two species, (ii)
statistically similar MAE for the other two species, and (iii)
considerably more precise forecasts for all species.

@@

@@

\begin{table}[ht]
\centering
\caption{Accuracy (mean absoluet error, MAE) and precision (90\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
are statisticially similar between models for \emph{P. secunda}
(\emph{P} = 0.0922).

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover; lower panels show the 90\% quantiles of predicted
proportional cover change. Mean predictions and their associated
uncertainties are shown in separate panels for visual clarity.}
\end{figure}
